---
title: "poisson_discrete_example"
author: "Sean Ammirati"
date: "July 29, 2018"
output: html_document
---
```{r}
source('./poisson_discrete_updater_functions.R')
```
#Example : Allergic Reactions 

Example: Consider that we believe that people in most counties will have a severe allergic reaction around 1.5 times per day. 
We have an expert who has just moved to XYZ county who thinks that this is not the case in the county. He wants to test this hypothesis on 
stretches of the last two weeks, where he believes we have seen a lower number of alergic reactions. 
(There was 12 incidents in the first 6 days, and zero in the last seven)

Suppose that we know (by some empirical estimation) that counties in the US approximately have probabilities .1,.2,.3,.2,.15,.05 of their rates 
being .5,1,1.5,2,2.5,4 respectively.
We believe that the distribution of the number of severe alergic reactions follows a Poisson distribution with rate lambda. That is, the rate is constant in 
a given county, and each event occurs independently of one another (i.e., one person having an alergic reaction does not effect the probability of another
having one). Given these assumptions, we want to check the probability of the rates given our new data. 

```{r}
poslam_example <- c(.5,1,1.5,2,2.5,4)
probs_example <- c(.1,.2,.3,.2,.15,.05)
X_example <- data.frame(k = c(12, 6), t = c(0, 7))

find_posterior(poslam_example, probs_example, X_example)
```
What is the conclusion? The expert seems to be on to something -- there appears to be more evidence at the current time that lambda is actually 1 rather than
1.5. More trials would find tune this result further. 