---
title: "Viterbi Algorithm Example"
author: "Sean Ammirati"
date: "August 4, 2018"
output: html_document
---


This investigates HMMs (Hidden Markov Models) with two possible states and six possible emissions using the viterbi algorithm.

Given a Markov process with two states (state 1 and state 2) and an arbitrary number of possible observed emissions , with transition probabilities
and emission probabilities given for each state, we can find the most likely sequence of states given some observed emissions. 
That is the goal of this exercise. 

This takes less time to run than considering all possible sequences of states -- we only need to consider the state probabilities forwards, rejecting 
paths that are unlikely at each point. 

The example which the following function defaults to is the case of a fair and loaded die, each with some probability of values 1-6. 
The roller does not show you what die he is using, but we know that he will generally switch between the two die with some transition probs.
The goal is then to find the most likely sequence of fair and loaded die given the observed sequence of emissions (for example 1,2,2,1,3).
```{r} 
source('./HMM_functions.R')
observed <- c(6,5,1,1,6,6,4,5,3,1,3,2,6,5,1,2,4,5,3,6,6,6,4,6,3,1,6,3,6,6,6,3,1,6,2,3,2,6,4,5,5,2,3,6,2,6,6,6,6,6,6,2,5,1,5,1,6,3,1)

results <- viterbialgorithm(observed)
resultswords <- c("Fair","Loaded")[results$path]
resultswords
```
So this tells us that given the observed die rolls, the most likely sequence of states is as shown above. 

This did not have to be done manually -- we can use the HMM package to do this. It will serve as a check. 

```{r}
require(HMM)
mod <- initHMM(c("Fair","Loaded"),c(1,2,3,4,5,6),startProbs = c(.5,.5),transProbs = rbind(c(.95,.05),c(.1,.90)),emissionProbs = rbind(c(rep(1/6,6)),c(rep(1/10,5),1/2)))
viterbi(mod,observed) == resultswords
## :) 
```